Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-12 NovelLBM.tex
First round complete

	\section{Introduction}
	For a solution to be well suited for usage as a bioink during bioprinting its rheology has to fulfill certain requirements. During the printing process it should flow easily through the nozzle, meaning the viscosity should be low. This is not only beneficial for ease of use, but is also crucial for survivability of the cells, as the stress they experience is dependent on the viscosity. However, after the printing process, it is desired, that the printed construct is stable. This can be temporarily archived by very high viscosities. Therefore, one approach to solve this is to use shear thinning fluids. Shear thinning fluids can easily be produced by dissolving polymers in water. This however, has the side effect of viscoelasticity. Due to the polymers the solution also exhibits elastic behavior. The influence of which on the printing process is not well understood yet. For realistic bioinks, the ratio of their zero shear viscosity and their infinite shear viscosity is large as is expected based on the desired attributes mentioned above. For such large viscosity ratios, simulating viscoelastic materials with Lattice Boltzmann methods was to our knowledge not yet possible.

	Such simulations would however be very beneficial. They allow the \glqq measurement\grqq{ }of parameters that are otherwise hard or even impossible to access. They allow the prediction of the outcome of experiments without performing them, which is especially helpful if the experiment cannot be easily performed. An example would be the design of new and improved printing nozzles. With simulations, the geometry can be optimized to a degree, that would be near impossible using iterative experiments. The large amounts of data simulations can provide allow better understanding of the fundamental processes, that cause observed behavior. \todo{"List some simulations for bionprinting, bioinks that have been done (restrict to only viscoelastic sims)"}

	Simulations using Lattice Boltzmann methods (LBM) are popular, as it has many advantages. It is very extensible and thus can deal with complex geometries, particles (e.g. cells) suspended in the fluid and many different interactions within and between those. Most importantly it is and easily parallelizable algorithm and thus can be run on GPUs or similar accelerators. This allows even large domains to be simulated in short amount of time with commodity GPUs and no need for any supercomputer. It is also easily possible to extend LBM to deal with viscoelastic fluids. This has been done before by multiple groups. There are many publications discussing such approaches in two dimensions\cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005}. Simulations in three dimensions have a higher computational cost and are not as abundant, but have also been performed by many\cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016}. Among these many are concerned about stability. Some include diffusivity of the polymer stress in their model to increase stability\cite{malaspinas2010,ma2020,wang2020,wang2021}. Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. However, none of these reach the relevant viscosity ratios for bioinks.

	Here we show our novel simulation approach for bioinks. It is specifically tailored to support arbitrary viscosity ratios and thus is able to simulate common bioinks or any other viscoelastic material with similar properties while providing high accuracy and stability. This is archived by subtracting a viscous stress from the polymer stress calculated from the PTT model. This offset is corrected by increasing the viscosity of the base LBM algorithm accordingly.

@@

Due to the complexity of viscoelastic fluids, there is an abundance of (semi-)empirical models to describe rheological data.
Common options include the Oldroyd-B \cite{oldroyd1958} and the FENE-P model \cite{bird1980,bird1987}.
The Oldroyd-B model is not appropriate for most real complex fluids as it features a shear-independent viscosity in clear contradiction to the experimental data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
Although the FENE-P does reproduce shear-thinning, we were not able to obtain a satisfactory fit to our data for alginate and methyl cellulose with physically reasonable values of the model parameters.
The best fit was obtained with the so-called PTT model due to Phan-Thien and Tanner \cite{phan-thien1977,phan-thien1978} which was for explicitly derived for polymer networks in solution and thus appears to capture well the essential physics behind the alginate solution. \SG{ check that these papers have the exponentail version }.

@@

For a simple shear flow in the $xy$-plane, an analytical expression for $ \eta $ and $N_1$ is given by \cite{ferras2019} which (with $\xi=0$) reads
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta _\mathrm{s} \ma{D}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right)
	\end{equation}
where $ \mathrm{Wi} = \lambda \dot{ \gamma  }$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }
\SG{ I did not find these equations in \cite{ferras2019}? }
Fitting these equations to the rheological data for alginate and methy cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $ \eta _ \mathrm{s} = 1 \si{ mPa } \cdot \si{ s }$ as the viscosity of the PBS buffer. \SG{ give ref }
At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent values $ \eta _ \mathrm{s}$.
The obtained parameters for alginate are
	\begin{align}
		\eta_\text{p} &= \SI{48.2\pm0.4}{\pascal\second}\\
		\lambda &= \SI{0.343 \pm 0.004}{\second}\\
		\epsilon &= \num{0.545 \pm 0.010}.
	\end{align}
while the parameters for the three concentrations of methyl cellulose are given in table~\ref{tab:MC_parameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c}
%			Concentration/\unit{\percent} & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ \\
			Concentration/ percent & $\eta_\text{p}/mPas$ & $\lambda/ms$ & $\epsilon$ \\
			\hline
			\num{0.49} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} \\
			\num{0.59} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} \\
			\num{0.83} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006}
		\end{tabular}
		\caption{Parameters of the PTT modelobtained when fitting the methyl cellulose solutions.}
		\label{tab:MC_parameters}
	\end{table}

@@

Since the early days of LBM methods for viscoelastic fluids have been developed. 
Keeping in mind our target geometries, we focus here on three dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a fluid obeying Jeffery's viscoelastic model \cite{lallemand2003}. \SG{ can we say more about Jeffery? Why is it bad? Is it shear-thinning? }
More advcanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells. 
The work of \cite{malaspinas2010} and \cite{ma2020} considered in addition the FENE-P model solving the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM while \cite{wang2020} used this approach for the Oldroyd-B model.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \SG{ cite }, which introduces a small error compared to the analytical solution of the constitutive equation. 
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to higher stability \cite{dzanic2022}.

@@

\caption{Covered ranges of $ \mathrm{Wi}$ and $ R_ \eta $ for existing 3D LB methods.}
\end{table}

@@

The key difficulty for simulating many realistic viscoelastic fluids is the unbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_ \eta $ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^4$ (see figure~\ref{fig:fit_alginate} and \ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polmer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} &= 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	
\end{align}
with the transfered (''shuffled'') stress
\begin{align}
	\ma{ \tau }_ \mathrm{shuffle} = 2 \alpha _ \mathrm{s} \eta _ \mathrm{p} \ma{D}
\end{align}
containing an arbitrary parameter $0 < \alpha _ \mathrm{s} < 1$.
Finally, this results in
\begin{align}
	\ma{ \sigma } &=  2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $ \alpha _ \mathrm{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity while from the polymer stress determined by equation~\ref{eq:polymerStress} the amount $\alpha_\text{s}\eta_\text{p} \ma{D}$ is subtracted at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and 0.49\% methyl cellulose solutions with a viscosity ratio $R_ \eta = 48 \cdot 10 ^3$ and $R_ \eta = 18.7$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with \SG{ XX } LBM nodes.
For the viscosity shuffling, we set $ \alpha _ \mathrm{s} = XX$.
Along the flow direction, the channel is periodic and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter}. \SG{ pressure for MC? }
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }.
After \SG{ XX } LBM time steps corresponding to \SG{ XX } of physical time, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
As shown in figure~\ref{fig:2DAlginateU}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
The spatially resolved errors are shown in figure~\ref{fig:2DAlginateErr} in the appendix.
Averaging over the channel diameter, we obtain an L2 error of \num{9e-4} and \SG{ XX } for alginate and methyl cellulose, respectively..

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in planar channel flow. \SG{use $y$ instead of $r$ in 2D}. 
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.}
		\label{fig:2DAlginateU}
	\end{figure}

We now proceed to evaluate the viscosity shuffling method for a cylindrical pipe flow.
The semi-analytical solution for this situation can be derived in a similar fashion as for the planar case and is given in appendix ~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical sitation is slightly more complex due to discretization errors (''staircase effect'') when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameter set as above, but double the pressure gradient to \SG{ XX }. \SG{ why? }
The channel has a diameter of \SG{ XX } and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure \ref{fig:3DAlginateU} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014}.
The spatially resolved error shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow. 
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.}
		\label{fig:3DAlginateU}
	\end{figure}

	%We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
	\subsection{Comparison with literature}
	Malaspinas et al.\cite{ma2020} provide a detailed validation and claim very high accuracy. Therefore, we pick this as a comparison. We aim at the reproduction of figures 17 and 18 from said paper. We calculate this 2D flow in 3D as the viscoelastic algorithm has not been validated for 2D velocity sets. Otherwise, the geometry is identical. It is not entirely clear to us, how they define typical scales. We define the typical velocity as the maximum velocity, as is likely done in the paper. The typical length, we set as half of the channel width. We define the typical shear rate as typical velocity divided by typical length. We set $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the paper only gives ratios, and it probably does not matter here. Nevertheless, our lattice units do not line up exactly with those from Malaspinas, as our viscosity is different due to the shuffling.
	%Beware equation 61 in the paper is wrong. It should have an 8 instead of the 4.
	The parameters from the paper ($R_\nu$, $Wi$ and $N$) are varied as in the paper. 
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/malaspinas2010.eps}
		\caption{Double logarithmic plot of the L2 error as defined in equation \ref{eq:l2error} from Malaspinas et al.\cite{malaspinas2010} (dots) compared to our algorithm (x) as a function of grid resolution.}
		\label{fig:malaspinas2010}
	\end{figure}
	As can be seen in figure \ref{fig:malaspinas2010}, our results notably do not depend on the Weissenberg Number. While our error decreases with N with a constant power, Malaspinas et al. see diminishing returns. While these results are excellent, the error in our implementation seemingly is still dependent on the viscosity ratio. As the error is still small using our alginate parameters (see section \ref{sec:channelFlow}), this effect is irrelevant. With this, one can see, that we are not only capable to simulate parameters not yet accessible, but our implementation is also better for the parameters used in literature. Knowing our simulations to be good we in the following show results for interesting systems, that are not analytically solvable.

>>>>

%%
%% Abstract
%%

\section*{Abstract}

The simulation of viscoelastic liquids using the Lattice-Boltzmann method in full three dimensions remains a formidable numerical challenge.
In particular the simulation of strongly shear-thinnig fluids, where the ratio between the high-shear and low-shear viscosities is large, is often prevented by stability problems.
Here we present a novel approach to overcome these issues.
The central idea is to artificially increase the solvent viscosity which allows the method to benefit from the very good stability properties of the LBM.
To compensate for this additional viscous stress, we reduce the polymer stress by the same amount.
We apply this novel method to simulate two realistic cell carrier liquids, methyl cellulose and alginate solutions, of which the latter exhibits a viscosity ratio of more than 10,000.

\clearpage

%%
%% Introduction
%%

\section{Introduction}

Microfluidic technoglogies involving living cells are becoming increasingly popular. 
Examples include medical diagnostics, where the pathological alteration of cell properties can serve as a marker for disease detection \cite{Toepfner.2018} or biofabrication where living cells are printed to create artificial tissues \cite{possl2021, Poologasundarampillai.2021, oconnell2022}. 
In all these applications cells are suspended in a cell carrier fluid which typically contains dissolved polymers making the fluid highly viscoelastic. 
The motivation behind using viscoelastic fluids differs between applications. 
In medical diagnostics, the viscous stress must be large enough to achieve detectable cell deformation, while in biofabrication the printed construct must be stable enough to retain its shape after the printing process. 
To satisfy the latter requirement, cell carriers used in biofabrication („bioinks“) typically exhibit a very high viscosity ratio with the zero-shear viscosity being many orders of magnitude larger than the pure solvent viscosity \cite{paxton2017}.

To understand the rheology of such complex liquids in non-trivial geometries, computer simulations are an invaluable tool. 
While a number of numerical techniques have been developed, the simulation of viscoelastic liquids remain an involved subject \cite{Alves.2020}, in particular for high viscosity ratios. 
A very popular simulation technique for Newtonian fluids is the Lattice Boltzmann method \cite{Kruger_book} due to its simplicity and high parallelizability. 
LBM has also been extended to viscoelastic situations, mainly in two \cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005} but occassionally also in three \cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016} dimensions.
To alleviate stability problems, some of these works include an artificial diffusivity of the polymer stress \cite{malaspinas2010,ma2020,wang2020,wang2021}. 
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. 
So far, however, none of these methods has been applied to very high viscosity ratio fluids which are often used as cell carrier fluids in lab experiments or technological applications.

To fill this gap, here we develop a new variant of viscoelastic LBM which allows the stable and accurate simulation of very high viscosity ratio fluids.
Our technique is based on the insight that stability issues in tradtional viscoelastic LBMs arise dure to an unbalance between the very large polymer stress and the much smaller viscous stress of the solvent. 
We overcome this issue by including („shuffling“) part of the polymer stress into the viscous part of the LBM algorithm.

%	For a solution to be well suited for usage as a bioink during bioprinting its rheology has to fulfill certain requirements. During the printing process it should flow easily through the nozzle, meaning the viscosity should be low. This is not only beneficial for ease of use, but is also crucial for survivability of the cells, as the stress they experience is dependent on the viscosity. However, after the printing process, it is desired, that the printed construct is stable. This can be temporarily archived by very high viscosities. Therefore, one approach to solve this is to use shear thinning fluids. Shear thinning fluids can easily be produced by dissolving polymers in water. This however, has the side effect of viscoelasticity. Due to the polymers the solution also exhibits elastic behavior. The influence of which on the printing process is not well understood yet. For realistic bioinks, the ratio of their zero shear viscosity and their infinite shear viscosity is large as is expected based on the desired attributes mentioned above. For such large viscosity ratios, simulating viscoelastic materials with Lattice Boltzmann methods was to our knowledge not yet possible.

%	Such simulations would however be very beneficial. They allow the \glqq measurement\grqq{ }of parameters that are otherwise hard or even impossible to access. They allow the prediction of the outcome of experiments without performing them, which is especially helpful if the experiment cannot be easily performed. An example would be the design of new and improved printing nozzles. With simulations, the geometry can be optimized to a degree, that would be near impossible using iterative experiments. The large amounts of data simulations can provide allow better understanding of the fundamental processes, that cause observed behavior. \todo{"List some simulations for bionprinting, bioinks that have been done (restrict to only viscoelastic sims)"}

%	Simulations using Lattice Boltzmann methods (LBM) are popular, as it has many advantages. It is very extensible and thus can deal with complex geometries, particles (e.g. cells) suspended in the fluid and many different interactions within and between those. Most importantly it is and easily parallelizable algorithm and thus can be run on GPUs or similar accelerators. This allows even large domains to be simulated in short amount of time with commodity GPUs and no need for any supercomputer. It is also easily possible to extend LBM to deal with viscoelastic fluids. This has been done before by multiple groups. There are many publications discussing such approaches in two dimensions\cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005}. 		Simulations in three dimensions have a higher computational cost and are not as abundant, but have also been performed by many\cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016}. Among these many are concerned about stability. Some include diffusivity of the polymer stress in their model to increase stability\cite{malaspinas2010,ma2020,wang2020,wang2021}. Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. However, none of these reach the relevant viscosity ratios for bioinks.

%	Here we show our novel simulation approach for bioinks. It is specifically tailored to support arbitrary viscosity ratios and thus is able to simulate common bioinks or any other viscoelastic material with similar properties while providing high accuracy and stability. This is archived by subtracting a viscous stress from the polymer stress calculated from the PTT model. This offset is corrected by increasing the viscosity of the base LBM algorithm accordingly.

@@

Due to the complexity of viscoelastic fluids, there is an abundance of (semi-)empirical models to describe rheological data.
Common options include the Oldroyd-B \cite{oldroyd1958} and the FENE-P model \cite{bird1980,bird1987}.
The Oldroyd-B model is not appropriate for most real complex fluids as it features a shear-independent viscosity in clear contradiction to the experimental data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
Although the FENE-P does reproduce shear-thinning, we were not able to obtain a satisfactory fit to our data for alginate and methyl cellulose with physically reasonable values of the model parameters.
The best fit was obtained with the so-called PTT model due to Phan-Thien and Tanner \cite{phan-thien1977,phan-thien1978} which was for explicitly derived for polymer networks in solution and thus appears to capture well the essential physics behind the alginate solution. \SG{ check that these papers have the exponential version }.

@@

For a simple shear flow in the $xy$-plane, an analytical expression for $ \eta $ and $N_1$ is given by \cite{ferras2019} which (with $\xi=0$) reads
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta _\mathrm{s} \ma{D} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $ \mathrm{Wi} = \lambda \dot{ \gamma  }$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }
\SG{ I did not find these equations in \cite{ferras2019}? }
Fitting these equations to the rheological data for alginate and methy cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $ \eta _ \mathrm{s} = 1 \si{ mPa } \cdot \si{ s }$ as the viscosity of the PBS buffer. \SG{ give ref }
At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent values $ \eta _ \mathrm{s}$.
The obtained parameters for alginate are
	\begin{align}
		\eta_\text{p} &= \SI{48.2\pm0.4}{\pascal\second}\\
		\lambda &= \SI{0.343 \pm 0.004}{\second}\\
		\epsilon &= \num{0.545 \pm 0.010}.
	\end{align}
while the parameters for the three concentrations of methyl cellulose are given in table~\ref{tab:MC_parameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c}
%			Concentration/\unit{\percent} & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ \\
			Concentration/ percent & $\eta_\text{p}/mPas$ & $\lambda/ms$ & $\epsilon$ \\
			\hline
			\num{0.49} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} \\
			\num{0.59} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} \\
			\num{0.83} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006}
		\end{tabular}
		\caption{Parameters of the PTT modelobtained when fitting the methyl cellulose solutions.}
		\label{tab:MC_parameters}
	\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }
Since the early days of LBM methods for viscoelastic fluids have been developed. 
Keeping in mind our target geometries, we focus here on three dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a fluid obeying Jeffery's viscoelastic model \cite{lallemand2003}. \SG{ can we say more about Jeffery? Why is it bad? Is it shear-thinning? }
More advcanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells. 
The work of \cite{malaspinas2010} and \cite{ma2020} considered in addition the FENE-P model solving the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM while \cite{wang2020} used this approach for the Oldroyd-B model.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \SG{ cite }, which introduces a small error compared to the analytical solution of the constitutive equation. 
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to higher stability \cite{dzanic2022}.

@@

\caption{Covered ranges of $ \mathrm{Wi}$ and $ R_ \eta $ for existing 3D LB methods.}
\label{tab:viscRatio}
\end{table}

@@

The key difficulty for simulating many realistic viscoelastic fluids is the unbalance between the viscous stress of the solvent and the polymer stress.
While the former is handled efficiently and reliably by the LBM streaming-collision scheme, the latter enters the LB equations as an additional source term.
The relative magnitude of the two terms is given by the viscosity ratio $R_ \eta $ as defined in equation~\ref{eq:R_eta} which for many situations is of the order of $10^2-10^4$ (see figure~\ref{fig:fit_alginate} and \ref{tab:bioinks}).
The key idea of our scheme is to artificially increase the LBM viscous stress while at the same time reducing the polmer stress by the same amount.
Starting from equation~\ref{eq:totalStress}, we write
\begin{align}
	\ma{\sigma} &= 2\eta_\text{s}\ma{D} + \ma{\tau}_\text{shuffle} + \ma{\tau} - \ma{\tau}_\text{shuffle}	 \label{eq:shuffleStress}
\end{align}
with the transfered (''shuffled'') stress
\begin{align}
	\ma{ \tau }_ \mathrm{shuffle} = 2 \alpha _ \mathrm{s} \eta _ \mathrm{p} \ma{D}
\end{align}
containing an arbitrary parameter $0 < \alpha _ \mathrm{s} < 1$.
Finally, this results in
\begin{align}
	\ma{ \sigma } &=  2\left(\eta_\text{s} + \alpha_\text{s}\eta_\text{p}\right)\ma{D} + \left(\ma{\tau} - 2\alpha_\text{s}\eta_\text{p}\ma{D}\right).
\end{align}
In practice, the shuffling parameter $ \alpha _ \mathrm{s}$ is chosen before the start of the simulation.
The LBM solver is run with $\eta_\text{s} + \alpha_\text{s}\eta_\text{p}$ as its viscosity while from the polymer stress determined by equation~\ref{eq:polymerStress} the amount $\alpha_\text{s}\eta_\text{p} \ma{D}$ is subtracted at every time step before it is coupled into the LBM.

@@

We start by simulating the alginate and 0.49\% methyl cellulose solutions with a viscosity ratio $R_ \eta = 48 \cdot 10 ^3$ and $R_ \eta = 18.7$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with \SG{ XX } LBM nodes.
For the viscosity shuffling, we set $ \alpha _ \mathrm{s} = XX$.
Along the flow direction, the channel is periodic and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter}. \SG{ pressure for MC? }
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched \SG{ give details }.
After \SG{ XX } LBM time steps corresponding to \SG{ XX } of physical time, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
As shown in figure~\ref{fig:2DAlginateU}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as \SG{ give correct equn. }

The spatially resolved errors are shown in figure~\ref{fig:2DAlginateErr} in the appendix.
Averaging over the channel diameter, we obtain an L2 error of \num{9e-4} and \SG{ XX } for alginate and methyl cellulose, respectively..

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in planar channel flow. \SG{use $y$ instead of $r$ in 2D}. 
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.
		\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include or put into Si }}
		\label{fig:2DAlginateU}
	\end{figure}

We now proceed to evaluate the viscosity shuffling method for a cylindrical pipe flow.
The semi-analytical solution for this situation can be derived in a similar fashion as for the planar case and is given in appendix ~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical sitation is slightly more complex due to discretization errors (''staircase effect'') when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameter set as above, but double the pressure gradient to \SG{ XX }. \SG{ why? }
The channel has a diameter of \SG{ XX } and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure \ref{fig:3DAlginateU} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014}.
The spatially resolved error shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
These conclusions remain robust even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:2DParameterStudyErr} of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow. 
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.
				\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include or put into Si }}
		\label{fig:3DAlginateU}
	\end{figure}

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we compare the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $ \xi  = \epsilon  = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.
In figure~\ref{fig:malaspinas2010} we show the L2 error as function of lattice resolution for different Weissenberg numbers $ \mathrm{Wi}$ and \SG{ XX } $R_ \nu $.
To match with \cite{malaspinas2010}, we defined these as 
\begin{align}
	XX
\end{align}
with \SG{ ... }. \SG{ comment whether these definitions are identical or different than the definitions above. }
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined slightly differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
\SG{ Beware equation 61 in the paper is wrong. It should have an 8 instead of the 4. Do you use equation 61? }

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/malaspinas2010.eps}
		\caption{L2 error as defined in equation \ref{eq:l2error} from Malaspinas et al.\cite{malaspinas2010} (dots) compared to our algorithm (x) as a function of grid resolution.
		Blue and green crosses are not visible as they overlap exactly with the orange and red crosses, respectively, showing that the accuracy of our method is. independent of $ \mathrm{Wi}$ in the studied parameter range.}
		\label{fig:malaspinas2010}
	\end{figure}

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_ \nu $ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \SG{ give number, show dashed line in plot }.
